---
title: "Fig 1"
author: "Katsuhiro Ito"
date: "2026-09-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(rtracklayer)
library(DESeq2)
library(ggnewscale)
library(ggrepel)
library(ggrastr)
library(org.Hs.eg.db)
library(clusterProfiler)


###clean environment###
rm(list = ls())

```

## Data load

```{r data load}
file <- list.files("./matrix/")
file

# old run

oldfile <- file %>% str_subset(pattern = "^Sample_LRS")
oldlist <- lapply(1:length(oldfile), FUN = function(x){
  dir <- paste("./matrix/", oldfile[x], sep = "")
  y <- read.table(dir, header = T)
  return(y)
})

olddf <- data.frame(oldlist[[1]]$expected_count)
olddf <- cbind(olddf, oldlist[[2]]$expected_count)
for(i in 3:length(oldfile)){
  olddf <- cbind(olddf, oldlist[[i]]$expected_count)
}

colnames(olddf) <- oldfile %>% str_remove_all(pattern = "Sample_|.rsem.genes.results")

olddf$gene_id <- oldlist[[1]]$gene_id

oldtpm <- data.frame(oldlist[[1]]$TPM)
for(i in 2:length(oldfile)){
  oldtpm <- cbind(oldtpm, oldlist[[i]]$TPM)
}

colnames(oldtpm) <- oldfile %>% str_remove_all(pattern = "Sample_|.rsem.genes.results")
oldtpm$gene_id <- oldlist[[1]]$gene_id



# 2nd run

newfile <- file %>% str_subset(pattern = "^Sample_.[0-9]")

newlist <- lapply(1:length(newfile), FUN = function(x){
  dir <- paste("./matrix/", newfile[x], sep = "")
  y <- read.table(dir, header = T)
  return(y)
})

newdf <- data.frame(newlist[[1]]$expected_count)
newdf <- cbind(newdf, newlist[[2]]$expected_count)
for(i in 3:length(newfile)){
  newdf <- cbind(newdf, newlist[[i]]$expected_count)
}

colnames(newdf) <- newfile %>% str_remove_all(pattern = ".rsem.genes.results")

newdf$gene_id <- newlist[[1]]$gene_id

newtpm <- data.frame(newlist[[1]]$TPM)
for(i in 2:length(newfile)){
  newtpm <- cbind(newtpm, newlist[[i]]$TPM)
}

colnames(newtpm) <- newfile %>% str_remove_all(pattern = ".rsem.genes.results")
newtpm$gene_id <- newlist[[1]]$gene_id


# merge files

gtf_data <- import("./references/GRCh38.gencode.collapsed.gtf")

gtf_df <- data.frame(geneid = as.vector(gtf_data$gene_id), genename = as.vector(gtf_data$gene_name))

gtf_df <- gtf_df　%>% distinct(geneid,genename)

table(duplicated(gtf_df$geneid))
#no duplicated ensemblids

table(duplicated(gtf_df$genename))
#1264 duplicated genenames

newdf<- newdf %>% left_join(gtf_df, by = c("gene_id" = "geneid"))
newtpm<- newtpm %>% left_join(gtf_df, by = c("gene_id" = "geneid"))

olddf<- olddf %>% left_join(gtf_df, by = c("gene_id" = "geneid"))
oldtpm<- oldtpm %>% left_join(gtf_df, by = c("gene_id" = "geneid"))

### merge data ###
setdiff(olddf$gene_id,newdf$gene_id)
setdiff(newdf$gene_id,olddf$gene_id)

### 27 genes were not detected in new dataset ###
olddf[olddf$gene_id %in% setdiff(olddf$gene_id,newdf$gene_id),]


mergeddf <- newdf %>% left_join(olddf, by = "gene_id")

mergeddf <- mergeddf %>% dplyr::select(starts_with("Sample"),starts_with("LRS"),genename = genename.x)

mergedtpm <- newtpm %>% left_join(oldtpm, by = "gene_id")

mergedtpm <- mergedtpm %>% dplyr::select(starts_with("Sample"),starts_with("LRS"),genename = genename.x)

### 1633 gene duplicated: sum counts ###
table(duplicated(mergeddf$genename))

mergeddf <- mergeddf %>% group_by(genename) %>% summarise(across(where(is.numeric), sum, na.rm = TRUE))

table(duplicated(mergeddf$genename))



```

## DESEQ


```{r deseq2}

mergedcoldata <- data.frame(condition = ifelse(str_detect(colnames(mergeddf)[2:33], pattern = "NT|WT"), "WT",
                                               ifelse(str_detect(colnames(mergeddf)[2:33], pattern = "HIF1a"), "HIF1a",
                                                      ifelse(str_detect(colnames(mergeddf)[2:33], pattern = "HIF2a"), "HIF2a",
                                                             ifelse(str_detect(colnames(mergeddf)[2:33], pattern = "Belz"), "Bel","NC")
                                                      )
                                               )
),
id = str_split(colnames(mergeddf)[2:33], pattern = "_|-", simplify = T)[,2],
batch = c(rep("new",16),rep("old",16))
)

mergedcoldata$sample <- colnames(mergeddf)[2:33]
mergedcoldata$condition <- factor(mergedcoldata$condition, levels = c("WT", "NC","Bel", "HIF1a", "HIF2a"))
rownames(mergedcoldata) <- mergedcoldata$sample




######## deseq

deseq.custom <- function(dfset, coldata, cond, paired = F, batch = F){
  if(length(cond) != 1 ){
    message("2 group should be provided with |")
    message("For example, A|B")
    exit
  }
  coldata.comparison <- coldata %>% filter(str_detect(condition, pattern = cond))
  dfset.comparison <- dfset %>% dplyr::select(coldata.comparison$sample)
  rownames(dfset.comparison) <- dfset$gene_id
  print(colnames(dfset.comparison))
  if(paired){
    if(batch){
      dds <- DESeqDataSetFromMatrix(round(dfset.comparison), colData = coldata.comparison, design = ~id + batch + condition)
    }else{
      dds <- DESeqDataSetFromMatrix(round(dfset.comparison), colData = coldata.comparison, design = ~id + condition)
    }
  }else{
    dds <- DESeqDataSetFromMatrix(round(dfset.comparison), colData = coldata.comparison, design = ~ condition)
  }
  dds <- DESeq(dds)
  res <- results(dds)
  res.dfset <- data.frame(res)
  res.dfset$gene <- rownames(res.dfset)
  
  res.dfset$genename <- dfset$genename
  
  return(res.dfset)
}


WTvsHIF1apairedfinal <- deseq.custom(mergeddf,coldata = mergedcoldata, cond = "HIF1a|WT", paired = T, batch = F)

WTvsHIF1apairedfinal <- WTvsHIF1apairedfinal %>% 
  mutate(color = case_when(padj < 0.05 & log2FoldChange > 1 ~ "red",
                           padj < 0.05 & log2FoldChange < -1 ~ "blue",
                           TRUE ~ "grey"),
         label = if_else(genename %in% c("VEGFA","PGK1","CXCR5","CXCR6"), genename,"")) 


### remove 38 and 42 according to KO data, see text
WTvsHIF2apairedfinal <- deseq.custom(mergeddf %>% select(-starts_with("LRS_42"), -starts_with("LRS_38")),coldata = mergedcoldata %>% dplyr::filter(id != 42 & id !=38), cond = "HIF2a|WT", paired = T, batch = F)

WTvsHIF2apairedfinal <- WTvsHIF2apairedfinal %>% 
  mutate(color = case_when(padj < 0.05 & log2FoldChange > 1 ~ "red",
                           padj < 0.05 & log2FoldChange < -1 ~ "blue",
                           TRUE ~ "grey"),
         label = if_else(genename %in% c("IFNG","IL2","TNF", "EPAS1"), genename,"")) 


```


### Figure 1


```{r Fig 1}
ggplot(WTvsHIF1apairedfinal, aes(x = log2FoldChange, y = -log10(padj))) + 
  rasterise(geom_point(aes(color = color),size = 0.2), dpi = 600) + 
  scale_color_manual(values = c("lightblue","lightgrey","lightpink")) +
  new_scale_color() +
  geom_label_repel(aes(label = label, color = color),size = 2, max.overlaps = 10000000) +
  scale_color_manual(values = c("blue","grey","red")) +
  geom_vline(xintercept = 1, linetype = 3) +
  geom_vline(xintercept = -1, linetype = 3) + 
  geom_hline(yintercept = -log10(0.05), linetype = 3) +
  scale_x_continuous(breaks = c(-6,-4,-2,2,4,6), limits = c(-7,7)) +
  scale_y_continuous(breaks = c(0,25,50,75), limits = c(0,100)) +
  labs(x = "Log2 Fold Change", y = "-log10 adjusted p value") + 
  theme_classic() +
  theme(axis.line = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "none",
        text = element_text(size = 6))

ggsave("WTvsHIF1aKO.png", width = 3, height = 2, units = "in")
ggsave("WTvsHIF1aKO.svg", width = 3, height = 2, units = "in")

ggplot(WTvsHIF2apairedfinal, aes(x = log2FoldChange, y = -log10(padj))) + 
  rasterise(geom_point(aes(color = color),size = 0.2), dpi = 600) + 
  scale_color_manual(values = c("lightblue","lightgrey","lightpink")) +
  new_scale_color() +
  geom_label_repel(aes(label = label, color = color),size = 2, max.overlaps = 10000000) +
  scale_color_manual(values = c("blue","grey","red")) +
  geom_vline(xintercept = 1, linetype = 3) +
  geom_vline(xintercept = -1, linetype = 3) + 
  geom_hline(yintercept = -log10(0.05), linetype = 3) +
  scale_x_continuous(breaks = c(-6,-4,-2,2,4,6), limits = c(-7,7)) +
  scale_y_continuous(breaks = c(0,2.5,5,7.5,10), limits = c(0,10)) +
  labs(x = "Log2 Fold Change", y = "-log10 adjusted p value") + 
  theme_classic() +
  theme(axis.line = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "none",
        text = element_text(size = 6))

ggsave("WTvsHIF2aKO.png", width = 3, height = 2, units = "in")
ggsave("WTvsHIF2aKO.svg", width = 3, height = 2, units = "in")
```

### Supplementary Figure 1

```{r}
############ gene list ########
hif <- read.table("./20260805hypoxiagenelist.txt")

WTvsHIF1apairedfinal <- WTvsHIF1apairedfinal %>% 
  mutate(color = case_when(padj < 0.05 & log2FoldChange > 1 ~ "red",
                           padj < 0.05 & log2FoldChange < -1 ~ "blue",
                           TRUE ~ "grey"),
         label = if_else(genename %in% hif$V1 & abs(padj) < 0.05 & abs(log2FoldChange) > 1, genename,"")) 


ggplot(WTvsHIF1apairedfinal, aes(x = log2FoldChange, y = -log10(padj))) + 
  rasterise(geom_point(aes(color = color),size = 0.2), dpi = 600) + 
  scale_color_manual(values = c("lightblue","lightgrey","lightpink")) +
  new_scale_color() +
  geom_label_repel(aes(label = label, color = color),size = 2, max.overlaps = 10000000) +
  scale_color_manual(values = c("blue","grey","red")) +
  geom_vline(xintercept = 1, linetype = 3) +
  geom_vline(xintercept = -1, linetype = 3) + 
  geom_hline(yintercept = -log10(0.05), linetype = 3) +
  scale_x_continuous(breaks = c(-6,-4,-2,2,4,6), limits = c(-7,7)) +
  scale_y_continuous(breaks = c(0,25,50,75), limits = c(0,100)) +
  labs(x = "Log2 Fold Change", y = "-log10 adjusted p value") + 
  theme_classic() +
  theme(axis.line = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "none",
        text = element_text(size = 6))

ggsave("WTvsHIF1aKO_Hypoxiagenes.png", width = 3, height = 2, units = "in")
ggsave("WTvsHIF1aKO_Hypoxiagenes.svg", width = 3, height = 2, units = "in")

ggsave("WTvsHIF1aKO_Hypoxiagenes_large.png", width = 6, height = 4, units = "in")
ggsave("WTvsHIF1aKO_Hypoxiagenes_large.svg", width = 6, height = 4, units = "in")


WTvsHIF2apairedfinal <- WTvsHIF2apairedfinal %>% 
  mutate(color = case_when(padj < 0.05 & log2FoldChange > 1 ~ "red",
                           padj < 0.05 & log2FoldChange < -1 ~ "blue",
                           TRUE ~ "grey"),
         label = if_else(genename %in% hif$V1 & abs(padj) < 0.05, genename,"")) 


ggplot(WTvsHIF2apairedfinal, aes(x = log2FoldChange, y = -log10(padj))) + 
  rasterise(geom_point(aes(color = color),size = 0.2), dpi = 600) + 
  scale_color_manual(values = c("lightblue","lightgrey","lightpink")) +
  new_scale_color() +
  geom_label_repel(aes(label = label, color = color),size = 2, max.overlaps = 10000000) +
  scale_color_manual(values = c("blue","grey","red")) +
  geom_vline(xintercept = 1, linetype = 3) +
  geom_vline(xintercept = -1, linetype = 3) + 
  geom_hline(yintercept = -log10(0.05), linetype = 3) +
  scale_x_continuous(breaks = c(-6,-4,-2,2,4,6), limits = c(-7,7)) +
  scale_y_continuous(breaks = c(0,2.5,5,7.5,10), limits = c(0,10)) +
  labs(x = "Log2 Fold Change", y = "-log10 adjusted p value") + 
  theme_classic() +
  theme(axis.line = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "none",
        text = element_text(size = 6))

ggsave("WTvsHIF2aKO_hypoxia.png", width = 3, height = 2, units = "in")
ggsave("WTvsHIF2aKO_hypoxia.svg", width = 3, height = 2, units = "in")



```

### save files
```{r save files}

write.csv(WTvsHIF1apairedfinal, file = "results/WTvsHIF1afinal.csv", quote = F,row.names = T,
          col.names = F)
write.csv(WTvsHIF2apairedfinal, file = "results/WTvsHIF2afinal.csv", quote = F,row.names = T,
          col.names = F)
write.csv(mergeddf, file = "results/finaldf.csv", quote = F,row.names = T,
          col.names = F)
```


### Enrichment analysis

```{r, enrichment}
res_hif1a_df <- read_csv("results/WTvsHIF1afinal.csv" )
res_hif2a_df <- read_csv("results/WTvsHIF2afinal.csv" )



# Add ENTREZID mapping to both dataframes
res_hif1a_df$ENTREZID <- mapIds(org.Hs.eg.db,
                                keys = res_hif1a_df$genename,
                                column = "ENTREZID",
                                keytype = "SYMBOL",
                                multiVals = "first")

res_hif2a_df$ENTREZID <- mapIds(org.Hs.eg.db,
                                keys = res_hif2a_df$genename,
                                column = "ENTREZID",
                                keytype = "SYMBOL",
                                multiVals = "first")


# Define thresholds
padj_threshold <- 0.05
#lfc_threshold <- 0
lfc_threshold <- 1

# HIF1a gene lists
hif1a_up_genes <- res_hif1a_df %>%
  dplyr::filter(padj < padj_threshold & log2FoldChange > lfc_threshold) %>%
  dplyr::pull(ENTREZID) %>%
  na.omit() %>%
  as.character()

hif1a_down_genes <-  res_hif1a_df %>%
  dplyr::filter(padj < padj_threshold & log2FoldChange < -lfc_threshold) %>%
  dplyr::pull(ENTREZID) %>%
  na.omit() %>%
  as.character()

# HIF2a gene lists
hif2a_up_genes <-  res_hif2a_df %>%
  dplyr::filter(padj < padj_threshold & log2FoldChange > lfc_threshold) %>%
  dplyr::pull(ENTREZID) %>%
  na.omit() %>%
  as.character()

hif2a_down_genes <-  res_hif2a_df %>%
  dplyr::filter(padj < padj_threshold & log2FoldChange < -lfc_threshold) %>%
  dplyr::pull(ENTREZID) %>%
  na.omit() %>%
  as.character()

# Universe
universe <-  res_hif1a_df %>%
  dplyr::pull(ENTREZID) %>%
  na.omit() %>%
  as.character()

message("Gene counts for enrichment:")
message(paste("HIF1a Up:", length(hif1a_up_genes)))
message(paste("HIF1a Down:", length(hif1a_down_genes)))
message(paste("HIF2a Up:", length(hif2a_up_genes)))
message(paste("HIF2a Down:", length(hif2a_down_genes)))

## HIF1a Pathway Analysis
# GO enrichment for HIF1a upregulated genes
hif1a_go_up <- NULL
if(length(hif1a_up_genes) > 5) {
  hif1a_go_up <- enrichGO(
    gene = hif1a_up_genes,
    universe = universe,
    keyType = "ENTREZID",
    OrgDb = org.Hs.eg.db,
    ont = "ALL",
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.2,
    readable = TRUE
  )
  
  if(nrow(as.data.frame(hif1a_go_up)) > 0) {
    p <- dotplot(hif1a_go_up, showCategory = 20) +
      ggtitle("HIF1a KO: GO Enrichment (Upregulated)") +
      theme(axis.text.y = element_text(size = 9))
    print(p)
  } else {
    message("No significant GO terms for HIF1a upregulated genes")
  }
}

# GO enrichment for HIF1a downregulated genes
hif1a_go_down <- NULL
if(length(hif1a_down_genes) > 5) {
  hif1a_go_down <- enrichGO(
    gene = hif1a_down_genes,
    universe = universe,
    keyType = "ENTREZID",
    OrgDb = org.Hs.eg.db,
    ont = "ALL",
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.2,
    readable = TRUE
  )
  
  if(nrow(as.data.frame(hif1a_go_down)) > 0) {
    p <- dotplot(hif1a_go_down, showCategory = 20) +
      ggtitle("HIF1a KO: GO Enrichment (Downregulated)") +
      theme(axis.text.y = element_text(size = 9))
    print(p)
  } else {
    message("No significant GO terms for HIF1a downregulated genes")
  }
}

# KEGG enrichment for HIF1a
hif1a_kegg_up <- NULL
if(length(hif1a_up_genes) > 5) {
  hif1a_kegg_up <- enrichKEGG(
    gene = hif1a_up_genes,
    universe = universe,
    organism = "hsa",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.2
  )
  
  if(nrow(as.data.frame(hif1a_kegg_up)) > 0) {
    hif1a_kegg_up <- setReadable(hif1a_kegg_up, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")
    p <- dotplot(hif1a_kegg_up, showCategory = 20) +
      ggtitle("HIF1a KO: KEGG Pathways (Upregulated)") +
      theme(axis.text.y = element_text(size = 9))
    print(p)
  } else {
    message("No significant KEGG pathways for HIF1a upregulated genes")
  }
}

hif1a_kegg_down <- NULL
if(length(hif1a_down_genes) > 5) {
  hif1a_kegg_down <- enrichKEGG(
    gene = hif1a_down_genes,
    universe = universe,
    organism = "hsa",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.2
  )
  
  if(nrow(as.data.frame(hif1a_kegg_down)) > 0) {
    hif1a_kegg_down <- setReadable(hif1a_kegg_down, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")
    p <- dotplot(hif1a_kegg_down, showCategory = 20) +
      ggtitle("HIF1a KO: KEGG Pathways (Downregulated)") +
      theme(axis.text.y = element_text(size = 9))
    print(p)
  } else {
    message("No significant KEGG pathways for HIF1a downregulated genes")
  }
}
## HIF2a Pathway Analysis

# GO enrichment for HIF2a upregulated genes
hif2a_go_up <- NULL
if(length(hif2a_up_genes) > 0) {
  hif2a_go_up <- enrichGO(
    gene = hif2a_up_genes,
    universe = universe,
    keyType = "ENTREZID",
    OrgDb = org.Hs.eg.db,
    ont = "ALL",
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.2,
    readable = TRUE
  )
  
  if(nrow(as.data.frame(hif2a_go_up)) > 0) {
    p <- dotplot(hif2a_go_up, showCategory = 20) +
      ggtitle("HIF2a KO: GO Enrichment (Upregulated)") +
      theme(axis.text.y = element_text(size = 9))
    print(p)
  } else {
    message("No significant GO terms for HIF2a upregulated genes")
  }
}

# GO enrichment for HIF2a downregulated genes
hif2a_go_down <- NULL
if(length(hif2a_down_genes) > 0) {
  hif2a_go_down <- enrichGO(
    gene = hif2a_down_genes,
    universe = universe,
    keyType = "ENTREZID",
    OrgDb = org.Hs.eg.db,
    ont = "ALL",
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.2,
    readable = TRUE
  )
  
  if(nrow(as.data.frame(hif2a_go_down)) > 0) {
    p <- dotplot(hif2a_go_down, showCategory = 20) +
      ggtitle("HIF2a KO: GO Enrichment (Downregulated)") +
      theme(axis.text.y = element_text(size = 9))
    print(p)
  } else {
    message("No significant GO terms for HIF2a downregulated genes")
  }
}

# KEGG enrichment for HIF2a
hif2a_kegg_up <- NULL
if(length(hif2a_up_genes) > 0) {
  hif2a_kegg_up <- enrichKEGG(
    gene = hif2a_up_genes,
    universe = universe,
    organism = "hsa",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.2
  )
  
  if(nrow(as.data.frame(hif2a_kegg_up)) > 0) {
    hif2a_kegg_up <- setReadable(hif2a_kegg_up, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")
    p <- dotplot(hif2a_kegg_up, showCategory = 20) +
      ggtitle("HIF2a KO: KEGG Pathways (Upregulated)") +
      theme(axis.text.y = element_text(size = 9))
    print(p)
  } else {
    message("No significant KEGG pathways for HIF2a upregulated genes")
  }
}

hif2a_kegg_down <- NULL
if(length(hif2a_down_genes) > 0) {
  hif2a_kegg_down <- enrichKEGG(
    gene = hif2a_down_genes,
    universe = universe,
    organism = "hsa",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.2
  )
  
  if(nrow(as.data.frame(hif2a_kegg_down)) > 0) {
    hif2a_kegg_down <- setReadable(hif2a_kegg_down, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")
    p <- dotplot(hif2a_kegg_down, showCategory = 20) +
      ggtitle("HIF2a KO: KEGG Pathways (Downregulated)") +
      theme(axis.text.y = element_text(size = 9))
    print(p)
  } else {
    message("No significant KEGG pathways for HIF2a downregulated genes")
  }
}

###save
write.csv(as.data.frame(hif1a_go_up), file = "results/hif1agoup.csv", quote = F)
write.csv(as.data.frame(hif1a_go_down), file = "results/hif1agodown.csv", quote = F)
write.csv(as.data.frame(hif2a_go_up), file = "results/hif2agoup.csv", quote = F)
write.csv(as.data.frame(hif2a_go_down), file = "results/hif2agodown.csv", quote = F)


write.csv(as.data.frame(hif1a_kegg_up), file = "results/hif1akeggup.csv", quote = F)
write.csv(as.data.frame(hif1a_kegg_down), file = "results/hif1akeggdown.csv", quote = F)
write.csv(as.data.frame(hif2a_kegg_up), file = "results/hif2akeggup.csv", quote = F)
write.csv(as.data.frame(hif2a_kegg_down), file = "results/hif2akeggdown.csv", quote = F)


# Function to create pathway bar plot
create_pathway_barplot <- function(data, title, n_pathways = 10, bar_color = "#2E86AB") {
  
  # tryCatch
  df <- tryCatch({
    data
  }, error = function(e) {
    return(NULL)
  })
  
  if (!length(df) > 0) {
    message("No enrichment data")
    return(NULL)
  }
  
  # Select top pathways and prepare data
  plot_data <- data %>%
    arrange(p.adjust) %>%
    slice_head(n = n_pathways) %>%
    mutate(
      neglog10_padj = -log10(p.adjust),
      # Wrap long descriptions for better display
      Description = str_wrap(Description, width = 50),
      # Order by significance for plotting
      Description = factor(Description, levels = rev(Description))
    )
  
  # Create bar plot
  p <- ggplot(plot_data, aes(x = neglog10_padj, y = Description)) +
    geom_bar(stat = "identity", fill = bar_color, alpha = 0.8) +
    geom_vline(xintercept = -log10(0.05), linetype = "dashed",
               color = "red", alpha = 0.5, linewidth = 0.5) +
    labs(
      title = title,
      x = "-Log10 (Adjusted P-value)",
      y = NULL
    ) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 13),
      axis.text.y = element_text(size = 9.5),
      axis.text.x = element_text(size = 10),
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_line(color = "grey90")
    ) +
    scale_x_continuous(expand = c(0, 0),
                       limits = c(0, max(plot_data$neglog10_padj) * 1.1))
  
  return(p)
}

# HIF1a downregulated pathways (top 10)
p_hif1a_path <- create_pathway_barplot(
  as.data.frame(hif1a_go_down),
  "HIF1α KO: Top 10 Downregulated GO Pathways",
  n_pathways = 10,
  bar_color = "#2E86AB"
)

print(p_hif1a_path)

ggsave("HIF1aKO_bar.svg", width = 10, height = 6, unit = "in")
ggsave("HIF1aKO_bar.png", width = 10, height = 6, unit = "in")

# HIF2a downregulated pathways (top 10)
p_hif2a_path <- create_pathway_barplot(
  as.data.frame(hif2a_go_down),
  "HIF2α KO: Top 10 Downregulated GO Pathways",
  n_pathways = 10,
  bar_color = "#A23B72"
)

print(p_hif2a_path)

ggsave("HIF2aKO_bar.svg", width = 10, height = 6, unit = "in")
ggsave("HIF2aKO_bar.png", width = 10, height = 6, unit = "in")



```

# sessioninfo

```{r}
sessionInfo()
```

Warning: Your system is mis-configured: ‘/etc/localtime’ is not a symlinkR version 4.4.1 (2024-06-14)
Platform: x86_64-pc-linux-gnu
Running under: Red Hat Enterprise Linux 9.6 (Plow)

Matrix products: default
BLAS/LAPACK: FlexiBLAS OPENBLAS;  LAPACK version 3.10.1

locale:
 [1] LC_CTYPE=C.UTF-8       LC_NUMERIC=C           LC_TIME=C.UTF-8        LC_COLLATE=C.UTF-8     LC_MONETARY=C.UTF-8    LC_MESSAGES=C.UTF-8    LC_PAPER=C.UTF-8      
 [8] LC_NAME=C              LC_ADDRESS=C           LC_TELEPHONE=C         LC_MEASUREMENT=C.UTF-8 LC_IDENTIFICATION=C   

time zone: NA
tzcode source: system (glibc)

attached base packages:
[1] stats4    stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] clusterProfiler_4.12.0      org.Hs.eg.db_3.19.1         AnnotationDbi_1.66.0        ggrastr_1.0.2               ggrepel_0.9.5               ggnewscale_0.4.10          
 [7] rtracklayer_1.64.0          lubridate_1.9.3             forcats_1.0.0               stringr_1.5.1               dplyr_1.2.1                 purrr_1.0.2                
[13] readr_2.1.5                 tidyr_1.3.2                 tibble_3.2.1                ggplot2_4.0.3               tidyverse_2.0.0             DESeq2_1.44.0              
[19] SummarizedExperiment_1.34.0 Biobase_2.64.0              MatrixGenerics_1.16.0       matrixStats_1.3.0           GenomicRanges_1.56.1        GenomeInfoDb_1.40.1        
[25] IRanges_2.38.0              S4Vectors_0.42.0            BiocGenerics_0.50.0        

loaded via a namespace (and not attached):
  [1] RColorBrewer_1.1-3       rstudioapi_0.16.0        jsonlite_1.8.8           magrittr_2.0.3           ggbeeswarm_0.7.2         farver_2.1.2             rmarkdown_2.27          
  [8] ragg_1.3.2               BiocIO_1.14.0            fs_1.6.4                 zlibbioc_1.50.0          vctrs_0.7.3              Cairo_1.6-2              Rsamtools_2.20.0        
 [15] memoise_2.0.1            RCurl_1.98-1.14          ggtree_3.12.0            htmltools_0.5.8.1        S4Arrays_1.4.1           curl_5.2.1               SparseArray_1.4.8       
 [22] gridGraphics_0.5-1       plyr_1.8.9               cachem_1.1.0             GenomicAlignments_1.40.0 igraph_2.0.3             lifecycle_1.0.5          pkgconfig_2.0.3         
 [29] Matrix_1.7-0             R6_2.5.1                 fastmap_1.2.0            gson_0.1.0               GenomeInfoDbData_1.2.12  digest_0.6.36            aplot_0.2.3             
 [36] enrichplot_1.24.0        colorspace_2.1-0         patchwork_1.2.0          textshaping_0.4.0        RSQLite_2.3.7            labeling_0.4.3           fansi_1.0.6             
 [43] timechange_0.3.0         httr_1.4.7               polyclip_1.10-6          abind_1.4-5              compiler_4.4.1           bit64_4.0.5              withr_3.0.0             
 [50] S7_0.2.2                 BiocParallel_1.38.0      viridis_0.6.5            DBI_1.2.3                ggforce_0.4.2            MASS_7.3-61              DelayedArray_0.30.1     
 [57] rjson_0.2.21             HDO.db_0.99.1            tools_4.4.1              vipor_0.4.7              beeswarm_0.4.0           ape_5.8                  scatterpie_0.2.3        
 [64] glue_1.7.0               restfulr_0.0.15          nlme_3.1-165             GOSemSim_2.30.0          grid_4.4.1               rsconnect_1.3.1          shadowtext_0.1.3        
 [71] reshape2_1.4.4           fgsea_1.30.0             generics_0.1.3           gtable_0.3.6             tzdb_0.4.0               data.table_1.15.4        hms_1.1.3               
 [78] tidygraph_1.3.1          utf8_1.2.4               XVector_0.44.0           pillar_1.9.0             vroom_1.6.5              yulab.utils_0.1.4        splines_4.4.1           
 [85] tweenr_2.0.3             treeio_1.28.0            lattice_0.22-6           bit_4.0.5                tidyselect_1.2.1         GO.db_3.19.1             locfit_1.5-9.10         
 [92] Biostrings_2.72.1        knitr_1.47               gridExtra_2.3            xfun_0.45                graphlayouts_1.1.1       stringi_1.8.4            UCSC.utils_1.0.0        
 [99] lazyeval_0.2.2           ggfun_0.1.5              yaml_2.3.8               evaluate_0.24.0          codetools_0.2-20         ggraph_2.2.1             qvalue_2.36.0           
[106] ggplotify_0.1.2          cli_3.6.3                systemfonts_1.1.0        dichromat_2.0-0.1        Rcpp_1.1.1-1.1           png_0.1-8                XML_3.99-0.16.1         
[113] parallel_4.4.1           blob_1.2.4               DOSE_3.30.1              bitops_1.0-7             viridisLite_0.4.2        tidytree_0.4.6           scales_1.4.0            
[120] crayon_1.5.3             rlang_1.3.0              cowplot_1.2.0            fastmatch_1.1-4          KEGGREST_1.44.1         
